Formalizing Stack Safety as a Security Property

### 2023-02-01 submission5/core.tex
Reorganizing the introduction to try to thread the needle right away.

The term {\em stack safety} is associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But, unlike ``the heap,'' the stack does not correspond to a single
high-level language concept, making a formal specification elusive.
Protection mechanisms typically lack precise specifications, relying instead
on informal descriptions and examples of bad behaviors that they prevent.
\apt{Can we connect stack safety more explicitly to a broader idea of ``subroutine safety''
  as discussed in the intro? Note that this also includes registers.}

When the prevented behaviors are straightforward, describing them may be sufficient
to justify an enforcement mechanism's effectiveness. Less intuitive enforcement
mechanisms instead leave us to ask: what security do we actually hope to gain,
and are we successful?\apt{For this audience, the need for a \emph{formal}
  characterization doesn't need to be belabored.}

In this paper we focus \apt{Is this still accurate?} on one such non-intuitive enforcement mechanism,
the ``lazy'' stack safety micro-policies proposed by
Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}. These permit functions
to write into one another's frames, but taint the result so that the frame's
owner cannot access it. No existing characterization of stack safety
captures this style of safety.

We propose a formal characterization of ``observable'' stack safety based on
concepts from language-based security. Stack safety is a
combination of an integrity property (``the private
state in each caller's stack frame is held invariant by the callee''),
two confidentiality properties (``the caller's
and callee's behavior are insensitive to one another's private
state,''), and a control flow property. By defining the properties
in terms of the observable behavior of the system, we expand the
space of supported enforcement mechanisms to include those that
allow certain (intuitively incorrect) behaviors but prevent them
from causing harm.

The stack interacts with a large number of language features that are often excluded
from discussion of stack safety. Our properties support a system with
both caller- and callee-saved registers, arguments passed on the stack,
tail-call elimination, [exceptions, and concurrency.] They are modular by
design, in order to be further extensible to other features and architectures.

We use these properties to validate Roessler and DeHon's micro-policies
via property-based random testing. Our testing successfully detects violations
in several broken variants, including Roessler and DeHon's original lazy policy.

>>>>

The term {\em stack safety} is associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But, unlike ``the heap,'' the stack does not correspond to a single
high-level language concept. Rather, it corresponds to the fundamental
abstraction of functions, in all of the forms that they can take.
The protean nature of the functional abstraction makes stack safety
difficult to specify.

We propose a formal characterization of stack safety based on
concepts from language-based security. Stack safety is decomposed
into an integrity property and a confidentiality property for each
of the caller and the callee, as well as a control-flow property.

Our motivating enforcement mechanism,
the ``lazy'' stack safety micro-policies proposed by
Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}, permit functions
to write into one another's frames, but taint the result so that the frame's
owner cannot access it. No existing characterization of stack safety
captures this style of safety. We capture it by defining the properties in
terms of the observable behavior of the system.

The stack interacts with a large number of language features that are often excluded
from discussion of stack safety. Our properties support a system with
both caller- and callee-saved registers, arguments passed on the stack,
tail-call elimination, \ifexceptions exceptions \fi. They are modular by
design, in order to be further extensible to other features.

We validate our properties by using them to distinguish correct implementations
of Roessler and DeHon's micro-policies from incorrect ones
via property-based random testing. Our testing successfully detects violations
in several broken variants, including Roessler and DeHon's original lazy policy.
A fixed version of the policy does pass our tests.

### 2023-02-01 submission5/intro-v2.tex
Reorganizing the introduction to try to thread the needle right away.

Indeed, the call stack is a perennial target for low-level attacks, leading to
dire consequences, from leakage and corruption of private stack data
to control-flow hijacking.
%
The most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read) or \#7 (CWE-416, Use After Free).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
  memory safety statistics, etc.}.
\sna{Trying to think how to make the case that these CWEs are about the stack, since they are
  primarily described as memory attacks.}\apt{Agree: the out-of-bounds ones are often on the stack, but the NULL pointer deref is not \rb{Indeed, removed the last one.}
Also, this story is very well known, and doesn't need belaboring.}
%
To foil such attacks, a menagerie of
software and hardware protections have been proposed,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}.
  \ifaftersubmission\apt{Mostly from
  nick; there could be more}\bcp{Yes, going back to MIT days---we should
  include several more of these, if only to give readers the impression that
this is a well-studied mechanism (so formalizing its protections is
useful).}
\fi

@@

Security mechanisms designed to prevent a certain attack can be brittle---successfully
eliminating the attack as it exists at the time of publication, but leaving room for attackers
to find new, similar attacks.

To mitigate this risk, mechanisms should aim to enforce \emph{formal properties}
of safe behavior that can be proven or tested. Such properties become the
\emph{specification} against which enforcement can be validated; even enforcement mechanisms that do
not fulfill a property benefit from the ability to articulate \emph{why} they fail.

But most such properties are described at the level of a high-level programming language.
The stack as an abstraction exists below the level of
any language that uses it, and at the level of the ABI
\sna{Do we need to unpack the acronym ABI?}, the specification of a well-behaved
stack is minimal. It is not concerned with such questions as whether a caller's
frame should be readable or writable to its callee, leaving those concerns
to be determined by a high-level language built atop the stack.

Specifying stack safety in full generality would not merely be a matter of handling
a wide range of architectural variations. The basic security behavior that we expect
of the stack varies based on calling convention, compiler behavior, and the features of
whichever programming language originated the code.
(Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across language semantics, and therefore
amenable to a very general approach.)

@@

In this work, we propose an explicit specification, built from first principles
in the hopes that it will be widely applicable, and demonstrate that it supports
an enforcement mechanism that does not yet have one.
We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety (see \cref{???}.)
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)), defined formally
in \cref{sec:props}.

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. This decomposition can be thought of as a step toward
a universal definition, in that the definitions in (3) capture fundamental
notions of security that can then be applied flexibly to other systems by varying
the model in (1) and (2).
%
\rb{FIXME: Contrast some universal definitions (or a parametric framework) with
  the concrete instance we study, again with its interest properly motivated.
  (Coming in key ideas?) No ``gold standard''?}
\apt{Yes, there are at least two different kinds of parameterization going on.
  One has to do with which properties we want to, or are able to, enforce; e.g.,
  for a language with tail calls or exceptions, we need a more complex definition of the WBCF property itself.
  Another has to do with how standard features, like argument passing, are implemented
  by particular hardware/ABIs -- which shouldn't show up in the property definition.
  We need to be clearer about this.}
\sna{In discussion above, I tried to carve out some separation there, but more may be needed.}

To demonstrate the utility of our formal characterization, we use these
properties to validate and improve an existing enforcement mechanism, the
{\em stack safety micro-policies} of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}, re-implemented
in the Coq proof assistant on top of a \rb{formal} RISC-V specification.  We
use QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based testing
tool for Coq, to generate random programs and check
that Roessler and DeHon's micro-policies correctly detect the ones that
attempt to violate one of our properties. Furthermore, we
check that the testing framework is able to generate counterexamples
that violate our properties but are \emph{not} halted by incorrect
variants of the enforcement mechanisms, including ones that we accidentally created
while implementating the micro-policy and ones that we
intentionally crafted to increase our confidence in the effectiveness
of our testing.
%
\rb{This and the following paragraph could start more generally (we need to
  answer the question of why these experiments, what is their interest), and
  probably motivate the use of PBT (experiment results below show there's
  benefit in them!) and, somewhere at least, talk about other points in the
  spectrum.}

>>>>

Indeed, the call stack is a perennial target for low-level attacks.
%
The most recent release of the authoritative CWE Top 25
Most Dangerous Software Weaknesses
list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
shows general classes of vulnerabilities closely related to the stack \rb{(though not exclusively)}
consistently at and near the top, e.g., \#1 (CWE-787, Out-of-bounds Write), \#5 (CWE-125, Out-of-bounds Read) or \#7 (CWE-416, Use After Free).
%
Moreover, many other types of faults involve some kind of control or data flow
using data stored in the stack \rb{cf. Eternal war in memory, Google/Microsoft
  memory safety statistics, etc.}.

The variety in attacks on the stack is mirrored by the 
software and hardware protections that aim to prevent them,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges+21},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18}---we discuss these in more
detail in \cref{XXX}. Few of these articulate any formal property that they aim to enforce,
with the exception of Skorstengaard and Georges. Some are fundamentally ill-suited for
formalization, as they aim to impede attackers but could not provide universal protection.

In this work, we propose such a property, built from first principles
in the hopes that it will be widely applicable, and demonstrate that it supports
an enforcement mechanism that does not yet have one.
We use the formal tools of language-based
security~\cite{sabelfeld2003language}, decomposing stack safety into
the {\em integrity} and {\em confidentiality} of the caller’s local state
and the callee's behavior during the callee's execution, plus the control-flow protection
of {\em well-bracketed control flow}~\cite{SkorstengaardSTKJFP} (\(\wbcf\)).
\ifexceptions
We also extend WBCF to a setting with exceptions,
in which there may be multiple points at which a callee can ``return.''
\fi

Our target for validation is the ``Lazy Tagging and Clearing'' \(LTC\) micropolicy of Roessler
and DeHon \cite{DBLP:conf/sp/RoesslerD18}. \(LTC\) is peculiar in that function activations
may write into one another's stack framess, as long as the data cannot be accessed afterward.
It is therefore too weak to be characterized by the state-of-the-art properties from
Georges et al. We give a characterization that does apply to it, and expand the features that
it supports.

@@

Security mechanisms can be brittle, successfully eliminating one attack while leaving room for
other, similar attacks. To avoid this, we seek formal properties of safe behavior that can be proven
or tested. Such properties become the specification against which enforcement can be validated;
even enforcement mechanisms that do not fulfill a property benefit from the ability to articulate
\emph{why} they fail.

But most such properties are described at the level of a high-level programming language.
The stack as an abstraction exists below the level of
any language that uses it, and at the level of the ABI, the specification of a well-behaved
stack is minimal. It is not concerned with such questions as whether a caller's
frame should be readable or writable to its callee, leaving those concerns
to be determined by a high-level language built atop the stack.
(Contrast Azevedo de Amorim et al.'s \cite{DBLP:conf/post/AmorimHP18} work on heap safety,
which is broadly consistent across language semantics, and therefore
amenable to a very general approach.)

@@

We are not attempting to create a universal definition of stack safety for all systems.
Rather, we define the stack safety of a single, fairly realistic system, decomposed into
(1) a lightweight model of the system and its security-relevant features, from which we
derive the integrity, confidentiality, and control-flow requirements of each call and
return, (2) the application of that model to a particular target system, and (3)
the formal criteria for integrity, confidentiality, and control flow
to obey the requirements. This decomposition can be thought of as a step toward
a universal definition, in that the definitions in (3) capture fundamental
notions of security that can then be applied flexibly to other systems by varying
the model in (1) and (2).
%
We treat stack safety as multiple properties, rather than one multi-faceted property, because
not all enforcement mechanisms attempt to implement all aspects of stack safety (see \cref{???}.)
%
This gives us a total of five core properties: \(\wbcf\),
caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)), defined formally
in \cref{sec:props}.

We demonstrate that our properties are general in that they apply easily to systems with various
features, and that they support ``al la carte'' enforcement based on the security needs of the
system.

To validate our formal characterization, we use these
properties to test and improve an existing enforcement mechanism, the
{\em stack safety micro-policies} of Roessler and DeHon~\cite{DBLP:conf/sp/RoesslerD18}, re-implemented
in the Coq proof assistant on top of a \rb{formal} RISC-V specification.  We
use QuickChick~\cite{Denes:VSL2014,Pierce:SF4}, a property-based testing
tool for Coq, to generate random programs and check
that Roessler and DeHon's micro-policies correctly detect the ones that
attempt to violate one of our properties. Furthermore, we
check that the testing framework is able to generate counterexamples
that violate our properties but are \emph{not} halted by incorrect
variants of the enforcement mechanisms, including ones that we accidentally created
while implementating the micro-policy and ones that we
intentionally crafted to increase our confidence in the effectiveness
of our testing.
%
\rb{This and the following paragraph could start more generally (we need to
  answer the question of why these experiments, what is their interest), and
  probably motivate the use of PBT (experiment results below show there's
  benefit in them!) and, somewhere at least, talk about other points in the
  spectrum.}
